Fasano--Franceschini Test: an Implementation of a 2-Dimensional Kolmogorov--Smirnov in \proglang{R}

### 2021-06-17 journalMS/article.tex
RB edits, penultimate

@@

% R -> Peacock test (computationally inificent)
% Python <- Implementation
% C <- Implementation
The Kolmogorov--Smirnov (KS) is a non--parametric, univariate
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample
case,  whether the two
samples come from the same underlying distribution). First derived by
Kolmogorov and Smirnov in a series of
papers~\citep{Kolmogorov1933,Kolmogorov1933a,Smirnov1936,Smirnov1937,Smirnov1939,Smirnov1944,Smirnov1948},
the one-sample KS test defines the distribution of the quantity
$D_{KS}$, the maximal absolute difference between the empirical
cumulative distribution function (CDF) of a set of values and a
reference probability distribution. Kolmogorov and Smirnov's key
insight was proving the distribution of $D_{KS}$ was independent of
the CDFs being tested. Thus, the test can effectively be used to
compare any univariate empirical data distribution to any continuous
univariate reference distribution. The two-sample KS test could further
be used to compare any two univariate empirical data distributions
against each other to determine if they are drawn from the same
underlying univariate distribution.

@@

Currently, only the Peacock implementation of the 2-D two-sample KS
test is available in \proglang{R}~\citep{R} with the
\pkg{Peacock.test} package via the \fct{peacock2} function, but this
has been shown to be markedly slower than the Fasano and Franceschini
algorithm~\citep{Lopes2007}. A \proglang{C} implementation of the
Fasano--Franceschini test is available in~\cite{numericalRecipes};
however, arguments have been made to the validity of the
implementation of the test not being
distribution-free~\citep{Babu2006}. Furthermore, the \proglang{C}
implementation statistical testing is based on a fit to Monte Carlo
Simulation and is only valid for significance levels $\alpha \lessapprox 0.20$,
which may present potential issues in some analysis.

Here we present the \pkg{fasano.franceschini.test} package as an
\proglang{R} implementation of the 2-D two-sample KS test described
by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides two improvements over
the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; and (ii) the package implements a bootstrapping procedure
for improved significance testing and mitigates the limitations of
the test brought noted by~\citep{Babu2006}.

@@

The Kolmogorov--Smirnov (KS) test is a non--parametric method for
determining whether a sample is consistent with a given probability
distribution~\citep{Stephens1992a}.
%While a comprehensive discussion pertaining the intricacies of the 1-D KS test can be found elsewhere~\citep{Stephens1992a}, we will present a brief overview as motivation for the basis and intuition of the 2-D KS test.
%
%As described previously, the 1-D KS test can be divided into two variants of the same statistical procedure:
%
%\begin{leftbar}
%\textbf{The One-Sample Test}
%\begin{itemize}
%\item Compares the data to analytical distribution.
%\end{itemize}
%\textbf{The Two-Sample Test}
%\begin{itemize}
%\item Compares two data samples to determine if they are drawn from the same underlying probability distribution.
%\end{itemize}
%\end{leftbar}
%
%In both cases, the KS test works as follows:
%
In one dimension, the Kolmogorov-Smirnov statistic ($D_{KS}$) is the
defined by the maximum absolute difference between the cumulative
density functions of the data and model (one--sample), or between the
two data sets (two--sample), as illustrated in
\textbf{Figure~\ref{fig:kstest1D}}.
%. In the following example, the orange and blue curve in the left panel represents the underlying distribution (pdf) for sample1 and sample2, respectively (\textbf{Figure \ref{fig:kstest1D}}). The eCDF is then defined by data drawn from these underlying distribution, and the D--stat is the computed as the largest between difference between the eCDF (black dotted line) in the right panel.

@@

%Kolmogorov and Smirnov work mathematically defined the distribution for $D_{KS}$ to be distribution free, and further derived a table of critical values for hypothesis testing. Kolmogorov and Smirnov work was later refined by Kendall and Stuart ~\citep{Kendall1946} to define the integrated probability distribution for large values of n  ($n \geq 80$), having the asymptotic expression:

In the large--sample limit ($n \geq 80$), it can be shown~\citep{Kendall1946} that $D_{KS}$ converges in distribution to
\begin{equation} \label{eq:1}
\Phi(\lambda) = 2 \sum_{k=1}^{\infty} -1^{k-1}e^{-2k^2\lambda^2} \,.
\end{equation}
%Given the maximum absolute difference ($D_{KS}$) and \textbf{Equation~\ref{eq:1}}, one can  compute a $p$-value for  hypothesis testing to determine if there is a significance difference between the cumulative density functions of the data and model (one-sample), or between the two data sets (two-sample).
%
%\textbf{For the 1-Sample case}, the $p$-value probability of a given D is defined by
In the one-sample case with a sample of size $n$, the $p$~value is given by
\begin{equation} \label{eq:2}
\mathbb{P}(D > observed) = \Phi ( D\sqrt{n})\,;
\end{equation}
in the two-sample case, the $p$~value is given by
%where $N$ is the number of data points sampled. A $p$-value $< 0.05$, indicates a significant difference between the samples.
%
%\textbf{In the 2-Sample case}, the $p$-value probability of a given D is defined by
\begin{equation} \label{eq:3}
\mathbb{P}(D > observed) = \Phi \left( D\sqrt{\frac{n_1n_2}{n_1+n_2}} \right)\,.
\end{equation}
%
where $n_1$ and $n_2$ are the number of observations in the first and second samples respectively.
%A $p$-value $< 0.05$, indicates a significant difference between the samples.

\subsection{Higher dimensional variations: Peacock Test (1983) and Fasano--Franceschini Test (1987)}
%Similar to the 1-D KS test, the obvious solution to extend the theory into higher dimensional would be to define the maximal cumulative difference between the two 2-D distributions. This however, presents an issue as the cumulative distribution function is not well defined in more than one dimension. In 2-D, there are 4 ways (3 independent ways) of defining the cumulative distribution, since the direction in which we define choose to order the points in the x and y are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}). In N dimensional space there are $2^{N}-1$ independent ways of defining the cumulative distribution function~\citep{Peacock1983}.
Extending the above to two or higher dimension is complicated by the
fact that CDFs are not well-defined in more than one dimension.  In
2-D, there are 4 ways (3 independent ways) of defining the cumulative
distribution, since the direction in which we order the $x$ and $y$
points are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}); more
generally, in $k$-dimensional space there are $2^{k}-1$ independent
ways of defining the cumulative distribution
function~\citep{Peacock1983}.

\begin{figure}[t!]
\centering
\includegraphics{CDF2Dissue}
\caption{\label{fig:kstest2Dissue} Four ways (3 independent) of defining the cumulative distribution for a given point in 2-D.  Here, the orange point $(X,Y)$ is chosen as the origin; the density of observations may be integrated as
$\mathbb{P}(x < X, y > Y)$ (A);
$\mathbb{P}(x < X \cup y < Y)$ (B);
$\mathbb{P}(x < X, y < Y)$ (C);
$\mathbb{P}(x > X, y > Y)$ (D).
%$P(x < X, y < Y) + P(x < X, y > Y) + P(x > X, y < Y) = 1 - P(x > X, y > Y)$
}
\end{figure}

\cite{Peacock1983} solved the higher dimensionality issue by defining
the 2-D test statistic as the largest difference between the empirical
and theoretical cumulative distributions, after taking all possible
ordering combinations into account. Peacock's test thus computes the
total integrated probability---i.e.\,fraction of data---tin each of
the four quadrants around all possible points in the data. For
example, for a given $n$ points in a two-dimensional space, the
cumulative distribution functions is calculated in the $4n^2$
quadrants of the plane defined by all pairs ($X_i$ ,$Y_j$), $X_i$ and
$Y_j$ being coordinates of any pairs of points in the given sample. By
ranging over all possible pairs of data points and quadrants, the
2-dimensional $D$ statistic is defined by the maximal difference of
the integrated probabilities between samples.

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered in each point of the given samples to compute the
cumulative distribution functions.  Rather than looking over all
$X_i, Y_j \,;\; i,j \in [1,n]$, Fasano and Franceschini only used the detected points
$X_i, Y_i \,;\; i \in [1,n]$.
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

%\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the points from \textit{other} sample
as the origins to compute the maximal $D_{FF,2}$ with origins from the
second sample. $D_{FF,1}$ and $D_{FF,2}$  are then averaged to compute
the overall $D_{FF}$ for hypothesis testing -- e.g.
$D_{FF}=(D_{FF,1}+D_{FF,2})/2$.

It may be that some points are tied with the $X$ and/or $Y$
coordinates of the origin, creating an ambiguity when computing the
fraction of points in each quadrant. Since the
test attempts to define the maximal difference of the cumulative
probabilities, a natural solution would be to treat a point that is
tied with the current $X$ and/or $Y$ coordinates of the origin as equally
likely to have been drawn from any of the tied quadrants.
Hence, any data point
sharing the same $X$ or $Y$ coordinate as the origin is evenly
distributed across the tied quadrants, with each of the two quadrants
recieving half a count. Any data point sharing the both the same $X$ and $Y$
coordinates as the current origin (including the origin itself) is evenly
distributed across all quadrants, with all four quadrants recieveing a
quarter count.

\subsection{Defining the Fasano and Franceschini null distribution}

Using Monte Carlo simulation, Fasano and Franceschini created a
look-up table of critical values of $D_{FF}$ as a function of
$D_{FF}$, the sample size, and the coefficient of correlation $r$.
\cite{numericalRecipes} later defined an approximate fit to the lookup table
as follows.
%
For the a single sample of size $n$,
\begin{equation} \label{eq:4}
\mathbb{P}(d_{FF} > D_{FF}) = \Phi \left( \frac{D_{FF}\sqrt{n}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{n})} \right) \, .
\end{equation}
where $\Phi$ is the same function defined in Eq~\ref{eq:1}.
The two sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
n = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined in the usual way as
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})}{\sqrt{\sum_{i}(X_i-\bar{X})^2}\sqrt{\sum_{i}(Y_i-\bar{Y})^2}}\, .
\end{equation}

@@

\subsection{Fasano--Franceschini Test Usage}

In their paper, Fasano and Franceschini use Monte Carlo simulation to
approximate the distribution of $D_{FF}$ as a function of the sample
size $n$ and the coefficient of correlation $r$. Notably, unlike the 1-D KS
test, the distribution of
$D_{FF}$ is \textit{not} completely independent of the shape of the 2-D distribution of the
underlying data, but depends on the correlations between the variables.
In the case where the variables $X$ and $Y$  are perfectly
correlated ($r = 1$), the 2-D distribution lies along a single line and thus
the 1-D KS test could be used; at the other extreme were $X$ and $Y$  are perfectly
uncorrelated ($r = 0$), the 2-D distribution is independent in the $X$
and $Y$ directions and one could apply the 1-D KS test on the marginal
distributions on the 2-D plane.
%
Results from Monte Carlo simulation support these expectaions, showing
that the distribution of $D$ is nearly identical for varying
distributions with the same correlation coefficient~\citep{Fasano1987}.
The approximation
by  \cite{numericalRecipes} (Eq~\ref{eq:4}--\ref{eq:5})
can be used to test the significance levels for the 2-D K-S test using
the following code:
%
\begin{CodeChunk}
\begin{CodeInput}
R> #set seed for reproducible example
R> set.seed(123)
R>
R> #create 2-D samples with the same underlying distributions
R> sample1Data <- data.frame(
R>  x = rnorm(n = 100, mean = 0, sd = 1),
R>  y = rnorm(n = 100, mean = 0, sd = 1)
R> )
R> sample2Data <- data.frame(
R>  x = rnorm(n = 100, mean = 0, sd = 1),
R>  y = rnorm(n = 100, mean = 0, sd = 1)
R> )
R>
R> fasano.franceschini.test(sample1Data,sample2Data)
\end{CodeInput}
\begin{CodeOutput}
      2-D Two-sample Kolmogorov-Smirnov Test

@@

\subsection{Bootstrap Version of the Fasano--Franceschini Test}

It has been noted that the approximation from  \citep{numericalRecipes} is only accurate when $n \gtrsim 20$ and the $p$-value is less than (more significant than) $\sim 0.2$ \citep{Babu2006}. While this inaccuracy still allows a simple rejection decision to be made at any $\alpha\leq0.2$, it is sometimes useful to quantify large $p$ more exactly (such as if one was to do a cross-study concordance analysis comparing $p$~values between studies as in~\cite{Ness-Cohn2020}), and to apply it to smaller datasets.
%
%\RBnote{I'm not sure what this next sentence is getting at; elaborate or cut. - ENC: removed the next line}
%Moreover, additional arguments have been made to the validity of the $>1$-D KS tests as not being distribution-free \citep{Babu2006}.
%
To address these limitations, one can bootstrap the significance levels for the particular multidimensional statistic directly from the particular data set under study. As Fasano and Franceschini's paper was originally released in 1987, this approach was unfeasible at scale.  Today, modern computers can rapidly compute a bootstrapped null distribution of $D_{FF}$ from the data to test significance.

The \pkg{fasano.franceschini.test} R package implements a parallelized
bootstrapping procedure. The marginal distribution from 2-dimensional
data set is resampled with replacement to generate randomized
2-dimensional data sets \code{nBootStrap} times. The frequency count
by quadrant is performed for each bootstrapped resampling as described
above to compute the $D_{FF}$. The observed $D_{FF}$ is then compared
to the distribution of bootstrapped $D_{FF}$ to compute a $p$~value.
The bootstrapped version of the Fasano--Franceschini test can be
implemented as follows (see \fct{fasano.franceschini.test} for further
source code details and implementation).

@@

\subsection{Computational efficency}
\RBnote{This is weird without any data.  Can you benchmark using your package without bootstrap; your package with bootstrap and varying cores; and Peacock.test package to illustrate scaling?}
The main distinction between the Peacock and Fasano--Franceschini tests is in computational efficiency, as
illustrated by the benchmarking studies in \cite{Lopes2007}.  Briefly,
%\begin{itemize}
%\item The Fasano--Franceschini test runs in $O(n^2)$ time, and is advantageous to the Peacock test as data set sizes scale.
%\item The Peacock test is more computationally expensive, and runs in $O(n^3)$ time.
%\end{itemize}
the Fasano--Franceschini test runs in $O(n^2)$ time, and is advantageous to the Peacock test as data set sizes scale, while the Peacock test is more computationally expensive, and runs in $O(n^3)$ time.
%
(See the \pkg{Peacock.test} package for \proglang{R} implementation of the Peacock test.)

@@

The \pkg{fasano.franceschini.test} package is an \proglang{R} implementation of the 2-D two-sample KS test as defined by Fasano and Franceschini~\citep{Fasano1987}.
It improves upon existing packages by implementing
  (i) a fast algorithm;
  (ii) a extension to the Fasano and Franceschini test for dealing with ties; and
  (iii) parallelized bootstrapping for improved statistical testing.
  See the complete package documentation and source code via the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/} or the package website at \url{https://nesscoder.github.io/fasano.franceschini.test/}.

>>>>

\hyphenation{Fa-sa-no=Fran-ces-chi-ni}

@@

The Kolmogorov--Smirnov (KS) is a non--parametric, univariate
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample
case,  whether the two
samples come from the same underlying distribution). First derived by
Kolmogorov and Smirnov in a series of
papers~\citep{Kolmogorov1933,Kolmogorov1933a,Smirnov1936,Smirnov1937,Smirnov1939,Smirnov1944,Smirnov1948},
the one-sample KS test defines the distribution of the quantity
$D_{KS}$, the maximal absolute difference between the empirical
cumulative distribution function (CDF) of a set of values and a
reference probability distribution. Kolmogorov and Smirnov's key
insight was proving the distribution of $D_{KS}$ was independent of
the CDFs being tested. Thus, the test can effectively be used to
compare any univariate empirical data distribution to any continuous
univariate reference distribution. The two-sample KS test could further
be used to compare any two univariate empirical data distributions
against each other to determine if they are drawn from the same
underlying univariate distribution.

@@

Currently, only the Peacock implementation of the 2-D two-sample KS
test is available in \proglang{R}~\citep{R} with the
\pkg{Peacock.test} package via the \fct{peacock2} function, but this
has been shown to be markedly slower than the Fasano and Franceschini
algorithm~\citep{Lopes2007}. A \proglang{C} implementation of the
Fasano--Franceschini test is available in~\cite{numericalRecipes};
however, arguments have been made to the validity of the
implementation of the test not being
distribution-free~\citep{Babu2006}. Furthermore, in the \proglang{C}
implementation, statistical testing is based on a fit to Monte Carlo
simulation that is only valid for significance levels $\alpha \lessapprox 0.20$.

Here we present the \pkg{fasano.franceschini.test} package as an
\proglang{R} implementation of the 2-D two-sample KS test described
by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides two improvements over
the current 2-D KS test available on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; and (ii) the package implements a bootstrapping procedure
for improved significance testing and mitigates the limitations of
the test brought noted by~\citet{Babu2006}.

@@

The Kolmogorov--Smirnov (KS) test is a non--parametric method for
determining whether a sample is consistent with a given probability
distribution~\citep{Stephens1992a}.
%
In one dimension, the Kolmogorov-Smirnov statistic ($D_{KS}$) is the
defined by the maximum absolute difference between the cumulative
density functions of the data and model (one--sample), or between the
two data sets (two--sample), as illustrated in
\textbf{Figure~\ref{fig:kstest1D}}.

@@

In the large--sample limit ($n \geq 80$), it can be
shown~\citep{Kendall1946} that $D_{KS}$ converges in distribution to
\begin{equation} \label{eq:1}
D_{KS} \overset{d}{\rightarrow} \Phi(\lambda) = 2 \sum_{k=1}^{\infty} -1^{k-1}e^{-2k^2\lambda^2} \,.
\end{equation}
In the one-sample case with a sample of size $n$, the $p$~value is given by
\begin{equation} \label{eq:2}
\mathbb{P}(D > observed) = \Phi ( D\sqrt{n})\,;
\end{equation}
in the two-sample case, the $p$~value is given by
\begin{equation} \label{eq:3}
\mathbb{P}(D > observed) = \Phi \left( D\sqrt{\frac{n_1n_2}{n_1+n_2}} \right)\,.
\end{equation}
where $n_1$ and $n_2$ are the number of observations in the first and second samples respectively.

\subsection{Higher dimensional variations: Peacock Test (1983) and Fasano--Franceschini Test (1987)}
Extending the above to two or higher dimension is complicated by the
fact that CDFs are not well-defined in more than one dimension.  In
2-D, there are 4 ways (3 independent) of defining the cumulative
distribution, since the direction in which we order the $x$ and $y$
points is arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}); more
generally, in $k$-dimensional space there are $2^{k}-1$ independent
ways of defining the cumulative distribution
function~\citep{Peacock1983}.

\begin{figure}[t!]
\centering
\includegraphics{CDF2Dissue}
\caption{\label{fig:kstest2Dissue} Four ways (3 independent) of defining the cumulative distribution for a given point in 2-D.  Here, the orange point $(X,Y)$ is chosen as the origin; the density of observations may be integrated as
$\mathbb{P}(x < X, y > Y)$ (A);
$\mathbb{P}(x < X \cup y < Y)$ (B);
$\mathbb{P}(x < X, y < Y)$ (C);
$\mathbb{P}(x > X, y > Y)$ (D).
}
\end{figure}

\cite{Peacock1983} solved the higher dimensionality issue by defining
the 2-D test statistic as the largest difference between the empirical
and theoretical cumulative distributions, after taking all possible
ordering combinations into account. Peacock's test thus computes the
total probability---i.e.\,fraction of data---in each of
the four quadrants around all possible tuples in the data. For
example, for $n$ points in a two-dimensional space, the empirical
cumulative distribution functions is calculated in the $4n^2$
quadrants of the plane defined by all pairs $(X_i, Y_j): i,j\in[1,n]$, where $X_i$ and
$Y_j$ are any observed value of $x$ and $y$ (whether or not they are observed as a pair).
There are $n^2$ such pairs, each of which can define four quadrants in the 2-D plane;
by ranging over all possible pairs of data points and quadrants, the
2-dimensional $D$ statistic is defined by the maximal difference of
the integrated probabilities between samples.

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered on each observed $(x, y)$ pair to compute the
cumulative distribution functions. That is, rather than looking over all $n^2$ points
${(X_i, Y_j): i,j \in [1,n]}$, Fasano and Franceschini only use the observed $n$ points
${(X_i, Y_i): i \in [1,n]}$.
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

%\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the points from \textit{other} sample
as the origins to compute the maximal $D_{FF,2}$ with origins from the
second sample. $D_{FF,1}$ and $D_{FF,2}$  are then averaged to compute
the overall $D_{FF}$ for hypothesis testing,
$D_{FF}=(D_{FF,1}+D_{FF,2})/2$.

It may be that some points are tied with the $X$ and/or $Y$
coordinates of the origin, creating an ambiguity when computing the
fraction of points in each quadrant. Since the
test attempts to define the maximal difference of the cumulative
probabilities, a natural solution would be to treat a point that is
tied with the current $X$ and/or $Y$ coordinates of the origin as equally
likely to have been drawn from any of the tied quadrants.
Hence, any data point
sharing the same $X$ or $Y$ coordinate as the origin is evenly
distributed across the tied quadrants, with each of the two quadrants
receiving half a count. Any data point sharing the both the same $X$ and $Y$
coordinates as the current origin (including the origin itself) is evenly
distributed across all quadrants, with all four quadrants receiving a
quarter count.

\subsection{Null distribution of $D_{FF}$}

Using Monte Carlo simulation, Fasano and Franceschini created a
look-up table of critical values of $D_{FF}$ as a function of
$D_{FF}$, the sample size, and the coefficient of correlation $r$.
\cite{numericalRecipes} later defined an approximate fit to the lookup table
as follows.
%
For a single sample of size $n$,
\begin{equation} \label{eq:4}
\mathbb{P}(d_{FF} > D_{FF}) = \Phi \left( \frac{D_{FF}\sqrt{n}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{n})} \right) \, .
\end{equation}
where $\Phi(\cdot)$ is as defined in Eq~\ref{eq:1}.
The two sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
n = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined in the usual way as
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})}{\sqrt{\sum_{i}(X_i-\bar{X})^2}\sqrt{\sum_{i}(Y_i-\bar{Y})^2}}\, .
\end{equation}

@@

\subsection{Fasano--Franceschini test usage}

In their paper, Fasano and Franceschini use Monte Carlo simulation to
approximate the distribution of $D_{FF}$ as a function of the sample
size $n$ and the coefficient of correlation $r$. Notably, unlike the 1-D KS
test, the distribution of
$D_{FF}$ is \textit{not} completely independent of the shape of the 2-D distribution of the
underlying data, but depends on the correlations between the variables.
In the case where the variables $X$ and $Y$  are perfectly
correlated ($r = 1$), the 2-D distribution lies along a single line and thus
the 1-D KS test could be used; at the other extreme were $X$ and $Y$  are perfectly
uncorrelated ($r = 0$), the 2-D distribution is independent in the $X$
and $Y$ directions and one could apply the 1-D KS test on the marginal
distributions.
%
Results from Monte Carlo simulation support these expectations, showing
that the distribution of $D$ is nearly identical for varying
distributions with the same correlation coefficient~\citep{Fasano1987}.
The approximation
by  \cite{numericalRecipes} (Eq~\ref{eq:4}--\ref{eq:5})
can be used to test the significance levels for the 2-D K-S test using
the following code:
%
\begin{CodeChunk}
\begin{CodeInput}
R> #set seed for reproducible example
R> set.seed(123)
R>
R> #create 2-D samples with the same underlying distributions
R> sample1Data <- data.frame(
R>  x = rnorm(n = 100, mean = 0, sd = 1),
R>  y = rnorm(n = 100, mean = 0, sd = 1)
R> )
R> sample2Data <- data.frame(
R>  x = rnorm(n = 100, mean = 0, sd = 1),
R>  y = rnorm(n = 100, mean = 0, sd = 1)
R> )
R>
R> fasano.franceschini.test(sample1Data,sample2Data)
\end{CodeInput}
\begin{CodeOutput}
      2-D Two-sample Kolmogorov-Smirnov Test

@@

\subsection{Bootstrap version of the Fasano--Franceschini test}

It has been noted that the approximation from
\citet{numericalRecipes} is only accurate when $n \gtrsim 20$ and the
$p$-value is less than (more significant than) $\sim 0.2$
\citep{Babu2006}. While this inaccuracy still allows a simple
rejection decision to be made at any $\alpha\leq0.2$, it is sometimes
useful to quantify large $p$ more exactly (such as if one was to do a
cross-study concordance analysis comparing $p$~values between studies
as in~\cite{Ness-Cohn2020}), and to apply it to smaller datasets.
%
To address these limitations, one can bootstrap the significance
levels for the particular multidimensional statistic directly from the
particular data set under study. As Fasano and Franceschini's paper
was originally released in 1987, this approach was unfeasible at
scale.  Today, modern computers can rapidly compute a bootstrapped
null distribution of $D_{FF}$ from the data to test significance.

The \pkg{fasano.franceschini.test} R package implements a parallelized
bootstrapping procedure. The marginal distribution from 2-dimensional
data set is resampled with replacement to generate randomized
2-dimensional data sets \code{nBootStrap} times. The frequency count
by quadrant is performed for each bootstrapped resampling as described
above to compute the $D_{FF}$. The observed $D_{FF}$ is then compared
to the distribution of bootstrapped $D_{FF}$ to compute a $p$~value.
The bootstrapped version of the Fasano--Franceschini test can be
run as follows (see \fct{fasano.franceschini.test} for further
source code details and implementation).

@@

\subsection{Computational efficency}

\begin{figure}[hbt]
\centering
\includegraphics{benchmark}
\caption{\label{fig:bmark} Computational efficiency benchmarks.
\textbf{A:} Runtime of the 
Fasano--Franceschini test relative to the Peacock test at four different sample sizes
($n=10, 100, 1000, 5000$).  Points represent the
average of 10 benchmark runs. \textbf{B:} Runtime of the
Fasano--Franceschini bootstrapping procedure for various sample sizes
($n= 10, 100, 1000, 5000$) as a function of the number of cores used. Within each panel,
lines are colored by the number of bootstrap iterations (no bootstrap, 10, 100,
1000). Points represent the average of 10 benchmark runs. Note the logarithmic $y$-axis
in (B).}
\end{figure}

To assess the computational efficiency, we benchmarked the package as follows.
\RBnote{Please explain the benchmark here; make note also of the specs of the machine that was used,
inc # cores, processor speed, memory, and OS.}
The main distinction between the Peacock and Fasano--Franceschini
tests is in computational efficiency, with  Fasano--Franceschini
scaling as $O(n^2)$ relative to Peacock's complexity of  $O(n^3)$
\citep{Lopes2007}.  Our benchmarks also show this advantage, as shown
in \textbf{Figure~\ref{fig:bmark}A}.  While the implentation of the
bootstrapping procedure increases runtime in comparison to the
approximate fit from  \citet{numericalRecipes}, parallelization of the
Fasano--Franceschini test shows a four-fold reduction in run time when
parallelized across 8 cores. \textbf{Figure~\ref{fig:bmark}B}.
% (See the \pkg{Peacock.test} package for \proglang{R} implementation of the Peacock test.)

@@

The \pkg{fasano.franceschini.test} package is an \proglang{R} implementation of the 2-D two-sample KS test as defined by Fasano and Franceschini~\citep{Fasano1987}.
It improves upon existing packages by implementing
  a fast algorithm and 
  a parallelized bootstrapping procedure for improved statistical testing.
  Complete package documentation and source code is available via the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/} and the package website at \url{https://nesscoder.github.io/fasano.franceschini.test/}.
